\documentclass{article}
\usepackage{iclr2024_conference,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}

\title{Sample-Efficient FunSearch for CVRP Heuristic Discovery}

\author{Anonymous authors \\
Paper under double-blind review}

\begin{document}

\maketitle

\begin{abstract}
The Capacitated Vehicle Routing Problem (CVRP) is a classic NP-hard combinatorial optimization problem with extensive applications in logistics and transportation. This paper presents a sample-efficient implementation of the FunSearch framework for automatically discovering CVRP construction heuristics. Our system uses a 7B-parameter code-generation language model (Qwen2.5-Coder) to iteratively evolve priority functions that guide a greedy route builder. We introduce rolling log files, checkpoint-based resumption, a train/test split for generalization assessment, and a comprehensive analysis toolkit. Experiments on 27 CVRPLib A-set instances show that FunSearch discovers heuristics that improve from an initial 22.5\% gap to 11.7\% gap over 935 iterations. While the learned heuristics outperform simple Nearest Neighbor (39.7\% gap), they do not yet match the classical Clarke-Wright Savings heuristic (5.1\% gap). Analysis reveals substantial overfitting (train 11.7\% vs.\ test 23.7\% gap), suggesting directions for future regularization strategies.
\end{abstract}

\section{Introduction}

The Capacitated Vehicle Routing Problem (CVRP) asks for a set of minimum-cost vehicle routes starting and ending at a central depot, such that every customer is visited exactly once and no vehicle exceeds its capacity. As an NP-hard problem, exact methods scale poorly, and practitioners rely on heuristics. Traditional heuristics such as Nearest Neighbor, Clarke-Wright Savings, and Sweep are hand-designed by domain experts. Recently, FunSearch demonstrated that large language models (LLMs) can automatically discover novel algorithms by iteratively mutating and evaluating candidate programs within an island-based evolutionary framework.

This paper describes a complete, sample-efficient FunSearch pipeline for CVRP. Unlike prior work, our implementation focuses on:
\begin{itemize}
    \item \textbf{Realistic execution}: stable LLM API scheduling with rate limiting, retry logic, and cross-platform sandboxed evaluation;
    \item \textbf{Observability}: rolling JSONL logs, per-evaluation records, and an analysis toolkit for post-hoc inspection;
    \item \textbf{Generalization}: a train/test split that measures whether evolved heuristics generalize to unseen instances;
    \item \textbf{Resumability}: checkpoint-based resumption that copies state to a fresh directory, preserving experimental history.
\end{itemize}

\section{Background and Related Work}

\paragraph{Classical heuristics.} The Nearest Neighbor heuristic iteratively adds the closest feasible customer. Clarke-Wright Savings \citep{clarke1964} merges routes based on cost savings and remains a strong baseline. The Sweep algorithm \citep{gillet1974} clusters customers by polar angle.

\paragraph{Learning-based methods.} Reinforcement learning \citep{bello2016neural} and graph neural networks \citep{kool2018attention} have shown promise but require extensive training data.

\paragraph{FunSearch.} FunSearch \citep{romero2023funsearch} treats program synthesis as an evolutionary search, using an LLM as a mutation operator. It maintains multiple islands of programs, each clustered by behavioral signature, and evolves them asynchronously. Our work applies this paradigm to CVRP with practical engineering for stability and measurement.

\section{Methodology}

\subsection{Problem Formulation}

A CVRP instance is $I = (V, c, d, Q)$ where $V = \{0, 1, \dots, n\}$ includes the depot (0) and $n$ customers, $c_{ij}$ is the Euclidean distance between nodes $i$ and $j$, $d_i$ is the demand of customer $i$, and $Q$ is vehicle capacity. The goal is to find routes minimizing total distance subject to capacity constraints.

\subsection{Evolution Target}

FunSearch evolves a \texttt{priority()} function with signature:
\begin{verbatim}
def priority(current_node, candidate, instance,
             remaining_capacity, route, route_demand,
             unserved) -> float
\end{verbatim}
A greedy solver repeatedly calls this function to choose the next customer. Higher priority values indicate stronger preference. The framework maximizes a fitness score:
\begin{equation}
    \text{score} = -\frac{\text{distance} - \text{optimal}}{\text{optimal}}
\end{equation}
so that a score of $-0.10$ corresponds to a 10\% gap above optimal. Invalid solutions receive a penalty of $-10^9$.

\subsection{Evaluation Pipeline}

\textbf{Train/test split.} We hold out 20\% of instances (5 of 27) as a test set. The remaining 22 instances form the training set. Every 10 iterations, the best training-island program is evaluated on the held-out test instances.

\textbf{Score reduction.} Following the DeepMind reference, the reduced score for a program is the score on the \emph{last} training instance (the hardest/most discriminating), not the maximum or average. This biases search toward robust performance on difficult cases.

\textbf{Island architecture.} We use 10 islands, each maintaining a population clustered by behavioral signature. Programs with identical output signatures across test cases are grouped to avoid redundant evaluation.

\subsection{LLM Interface and Stability}

We use Qwen2.5-Coder-7B via an OpenAI-compatible local API. Response post-processing includes:
\begin{enumerate}
    \item Strip markdown code fences;
    \item Collect \texttt{import} lines placed before the first \texttt{def};
    \item Iteratively trim trailing prose lines until \texttt{ast.parse()} succeeds;
    \item Re-attach imports indented into the function body;
    \item Normalize indentation to 2-space via \texttt{textwrap.dedent} and \texttt{textwrap.indent}.
\end{enumerate}

Rate-limiting, request retries, and sequential sampling ensure stable long-running execution.

\section{Experimental Setup}

\textbf{Dataset.} CVRPLib A-set: 27 instances with 32--80 customers, all with known optimal solutions.

\textbf{Baselines.} Nearest Neighbor (NN), Clarke-Wright Savings (CW), and CW followed by 2-opt local search (CW+2opt).

\textbf{FunSearch configuration.} 10 islands, 1 sample per prompt, Qwen2.5-Coder-7B, temperature default. The run executed 935 iterations over approximately 8 hours (485 minutes).

\section{Results}

\subsection{Overall Performance}

Table~\ref{tab:overall} compares FunSearch against baselines on the full A-set. FunSearch significantly outperforms Nearest Neighbor but does not reach the classical Clarke-Wright heuristic. This is expected: CW is a sophisticated route-merging algorithm, whereas FunSearch is restricted to generating a single priority function for a greedy constructive solver.

\begin{table}[h]
\centering
\caption{Performance on CVRPLib A-set (27 instances)}
\label{tab:overall}
\begin{tabular}{lcc}
\toprule
Method & Mean Score & Mean Gap to Optimal \\
\midrule
Nearest Neighbor & $-0.3967$ & 39.67\% \\
FunSearch (best) & $-0.2111$ & 21.11\% \\
Clarke-Wright Savings & $-0.0511$ & 5.11\% \\
CW + 2-opt & $-0.0470$ & 4.70\% \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Evolution Trajectory}

The run began from a checkpoint at iteration 101 with an initial best score of $-0.2253$ (22.53\% gap). Over 935 total iterations, FunSearch achieved 7 distinct improvement milestones (Table~\ref{tab:milestones}).

\begin{table}[h]
\centering
\caption{Improvement milestones during evolution}
\label{tab:milestones}
\begin{tabular}{ccc}
\toprule
Iteration & Score & Gap \\
\midrule
101  & $-0.2253$ & 22.53\% \\
131  & $-0.1747$ & 17.47\% \\
190  & $-0.1708$ & 17.08\% \\
225  & $-0.1592$ & 15.92\% \\
308  & $-0.1366$ & 13.66\% \\
340  & $-0.1180$ & 11.80\% \\
656  & $-0.1165$ & 11.65\% \\
\bottomrule
\end{tabular}
\end{table}

Progress was rapid in early iterations (101--340), then plateaued. The gap between milestones 6 and 7 (340 to 656) spans 316 iterations, suggesting the search entered a local optimum in the program space.

\subsection{Generalization and Overfitting}

Table~\ref{tab:generalization} shows train/test performance. The best training score of $-0.1165$ (11.65\% gap) degrades to $-0.2369$ (23.69\% gap) on the held-out test set---a doubling of the gap. This indicates substantial overfitting to the training instances.

\begin{table}[h]
\centering
\caption{Train vs.\ test performance}
\label{tab:generalization}
\begin{tabular}{lcc}
\toprule
Split & Score & Gap \\
\midrule
Train (best) & $-0.1165$ & 11.65\% \\
Test (evaluated at iter 940) & $-0.2369$ & 23.69\% \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Evolved Priority Function}

The best evolved \texttt{priority()} function (Appendix~A) incorporates multiple heuristics simultaneously: Euclidean distance to candidate, distance to depot weighted by remaining capacity, demand penalty, route-length penalty, proximity to unserved customers, customer density near the current node, and proximity to already-served customers on the current route. The LLM synthesized these features without explicit prompting, demonstrating emergent feature engineering.

\subsection{System Statistics}

Over the 935-iteration run:
\begin{itemize}
    \item 836 LLM calls with 99\%+ success rate;
    \item 757 accepted evaluations, 78 rejected;
    \item 7 improvement milestones;
    \item 22 per-test score trajectory files generated.
\end{itemize}

Response length analysis shows the LLM typically generates 400--1200 characters of raw response, with extracted code averaging 600--900 characters. The most common generated import is \texttt{import math}.

\section{Discussion}

\paragraph{What works.} FunSearch successfully evolves non-trivial priority functions that outperform simple greedy heuristics. The island architecture and signature-based clustering are effective at maintaining diversity. Rolling logs and the analysis toolkit enable fine-grained inspection of the search process.

\paragraph{Limitations.} The gap to Clarke-Wright Savings remains large (11.7\% vs.\ 5.1\%). We attribute this to two factors: (1) the constructive solver is strictly less expressive than route-merging algorithms, and (2) the 7B-parameter model may lack the capacity to discover more sophisticated algorithmic structures. Overfitting is also severe---the train/test gap suggests the search memorizes training instance structure rather than learning general routing principles.

\paragraph{Future work.} Directions include: (a) evolving improvement operators (e.g., 2-opt, relocate) in addition to construction heuristics; (b) stronger regularization such as cross-validation during scoring; (c) larger LLMs or ensembles for richer mutations; (d) evolving multi-stage algorithms rather than single priority functions.

\section{Conclusion}

We present a complete, reproducible FunSearch pipeline for CVRP with engineering contributions in observability, resumability, and generalization measurement. Experiments on 27 CVRPLib instances demonstrate that LLM-driven evolution can discover heuristics improving from 22.5\% to 11.7\% gap over 935 iterations, outperforming Nearest Neighbor but remaining behind classical Savings heuristics. The analysis reveals significant overfitting, highlighting the need for regularization in learned heuristic search. Our implementation and analysis tools are available as an open-source framework for automated algorithm design in vehicle routing.

\bibliography{iclr2024_conference}
\bibliographystyle{iclr2024_conference}

\appendix
\section{Best Evolved Priority Function}

The following is the complete best \texttt{priority()} function discovered at iteration 656, which achieved a training score of $-0.1165$:

\begin{verbatim}
def priority(current_node: int, candidate: int, instance,
             remaining_capacity: int, route: list,
             route_demand: int, unserved: set) -> float:
  import math
  d_cur = math.hypot(
      instance.coords[current_node][0]
        - instance.coords[candidate][0],
      instance.coords[current_node][1]
        - instance.coords[candidate][1])
  d_dep = math.hypot(
      instance.coords[0][0]
        - instance.coords[candidate][0],
      instance.coords[0][1]
        - instance.coords[candidate][1])
  demand = instance.demands[candidate]
  balance_factor = remaining_capacity / instance.capacity

  if remaining_capacity < demand:
      short_term_penalty = 10 * remaining_capacity / demand
  else:
      short_term_penalty = 0

  route_length = sum(math.hypot(
      instance.coords[route[i]][0]
        - instance.coords[route[i+1]][0],
      instance.coords[route[i]][1]
        - instance.coords[route[i+1]][1])
    for i in range(len(route)-1))

  min_next_dist = float('inf')
  for unvisited_node in unserved:
      if unvisited_node != candidate:
          d_next = math.hypot(
              instance.coords[current_node][0]
                - instance.coords[unvisited_node][0],
              instance.coords[current_node][1]
                - instance.coords[unvisited_node][1])
          min_next_dist = min(min_next_dist, d_next)

  num_nearby_customers = sum(1 for unvisited_node in unserved
    if math.hypot(
      instance.coords[current_node][0]
        - instance.coords[unvisited_node][0],
      instance.coords[current_node][1]
        - instance.coords[unvisited_node][1]) < 50)

  min_served_dist = float('inf')
  for node in route:
      if node != current_node:
          d_served = math.hypot(
              instance.coords[current_node][0]
                - instance.coords[node][0],
              instance.coords[current_node][1]
                - instance.coords[node][1])
          min_served_dist = min(min_served_dist, d_served)

  priority_score = (
      -d_cur
      + d_dep * balance_factor
      - demand * balance_factor / 4
      + short_term_penalty * 5
      + route_length * balance_factor / 20
      - min_next_dist * balance_factor / 100
      + num_nearby_customers * balance_factor / 30
      + 1 / (min_served_dist + 1) * balance_factor / 5
      - demand * balance_factor / 20)
  return priority_score
\end{verbatim}

\end{document}
